Thesis Proposal

### 2019-10-10 ThesisProposal.Rmd
Revised after feedback Gerko on lit rev

---
title: "Thesis Proposal"
author: "Hanne Oberman (4216318)"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  #word_document: default
  pdf_document:
    fig_caption: yes
subtitle: Methodology and Statistics for the Behavioural, Biomedical and Social Sciences
bibliography: ThesisProposal.bib
csl: acm.csl

@@

Word count: 739

@@

With MI, many assumptions are made about the nature of the observed and missing parts of the data **[remove "and their relation to the 'true' data generating mode"?l** [@buur18]. Without proper evaluation of the imputations and the underlying assumptions, any drawn inference may erroneously be deemed valid. Such evaluation measures are currently missing or under-developed in MI software, like the world leading `R` package `MICE` [@mice11]. Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.

@@

The validity of MI data depends on numerous assumptions that--by definition--cannot be verified from the incomplete data. Consequently, existing evaluation methods rely on proxy measures. For the following assumptions, no reliable proxy measures have been proposed and/or implemented: 1) *ignorability* of the *missingness mechanism*; 2) suitability of the *imputation models*; and 3) *compatibility* with the MI algorithm. 
<!-- Rubin (1987b, 160-66) subdivided the work needed to create imputations into three tasks. The modeling task chooses a specific model for the data, the estimation task formulates the posterior parameters distribution given the model and the imputation task takes a random draws for the missing data by drawing successively from parameter and data distributions @buur18 par 4.5.1 -->

@@

1. A missingness mechanism is said to be ignorable when the 'cause' of the missingness does not depend on the missing data itself [@rubin87]. Evaluation of this assumption is usually done with sensitivity analyses to assess the robustness of inferences to violations. Some practical guidelines exist (e.g., [@nguy17]), but current MI software does not facilitate this methodology for empirical researchers.

@@

2. Suitable imputation models capture the relations between observed and missing parts of the data. These relations can be evaluated by plotting conditional distributions [@abay08]. Such visualizations are available in `MICE`, but subsequent statistical tests to quantify the relations with covariates are not yet provided. Additionally, there is potential to assess model fit by means of *over-imputation* [@buur18] or *double robustness* [@bang05].

@@

3. The third assumption is met when the MI algorithm converges to a stable distribution. However, conventional measures to diagnose convergence--e.g., Gelman and Rubin's [-@gelm92] statistic $\widehat{R}$--do not suit multiply imputed data [@lace07]. Therefore, empirical researchers have to rely on a visual inspection procedure that is theoretically equivalent to $\widehat{R}$ [@whit11]. Not only is visually assessing convergence difficult for the untrained eye, it might also be futile. The convergence properties of MI algorithms lack scientific consensus [@taka17], and some default `MICE` techniques might not converge to stable distributions at all [@murr18]. Moreover, convergence diagnostics for MI methods have not been systematically studied [@buur18].

In short, the existing literature provides both possibilities and limitations to evaluating the validity of multiply imputed data. The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods, and implement these in an interactive evaluation device for multiple imputation. This tool will aid applied researchers in drawing valid inference from incomplete datasets.

>>>>

---
title: "Thesis Proposal"
author: "Hanne Oberman (4216318)"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  #word_document: 
  pdf_document: 
    fig_caption: yes
subtitle: Methodology and Statistics for the Behavioural, Biomedical and Social Sciences
bibliography: ThesisProposal.bib
csl: acm.csl

@@

Word count: 713

@@

With MI, many assumptions are made about the nature of the observed and missing parts of the data and their relation to the 'true' data generating model [@buur18]. Without proper evaluation of the imputations and the underlying assumptions, any drawn inference may erroneously be deemed valid. Such evaluation measures are currently missing or under-developed in MI software, like the world leading `R` package `MICE` [@mice11]. Therefore, I will answer the following question: 'Which measures are vital for evaluating the validity of multiply imputed data?'.

@@

The validity of the MI solution depends on numerous assumptions that cannot be verified from the observed data alone. Consequently, existing evaluation methods rely on **[rephrase: proxy measures]**. For the following assumptions, no reliable *proxy measures* have been proposed and/or implemented: 1) *ignorability* of the *missingness mechanism*; 2) congruity of the *imputation models*; and 3) *compatibility* of the MI modeling procedure. 
<!-- Rubin (1987b, 160-66) subdivided the work needed to create imputations into three tasks. The modeling task chooses a specific model for the data, the estimation task formulates the posterior parameters distribution given the model and the imputation task takes a random draws for the missing data by drawing successively from parameter and data distributions @buur18 par 4.5.1 -->

@@

1. A missingness mechanism is said to be ignorable when the sprobability to be missing does not depend on the missing data itself [@rubin87]. Evaluation of this assumption is usually done with sensitivity analyses to assess the robustness of inferences to violations. Some practical guidelines exist (e.g., [@nguy17]), but current MI software does not facilitate this methodology for empirical researchers.

@@

2. Congruent imputation models capture all required relations between observed and missing parts of the data. The exent to which this has been successful can be evaluated by plotting conditional distributions [@abay08]. Such visualizations are available in `MICE`, but subsequent statistical tests to quantify the relations with covariates are not provided. Additionally, there is potential to assess model fit by means of *over-imputation* [@buur18] or *double robustness* [@bang05].

@@

3. The third assumption is met when the MI algorithm converges to a stable distribution. However, conventional measures to diagnose convergence-- e.g., Gelman and Rubin's [-@gelm92] statistic $\widehat{R}$ --are not applicable on multiply imputed data [@lace07]. Therefore, empirical researchers have to rely on visual inspection procedures that are theoretically equivalent to $\widehat{R}$ [@whit11]. Visually assessing convergence is not only difficult to the untrained eye, it might also be futile. The convergence properties of MI algorithms lack scientific consensus [@taka17], and some default `MICE` techniques might not converge to stable distributions at all [@murr18]. Moreover, convergence diagnostics for MI methods have not been systematically studied [@buur18].

In short, the existing literature provides both possibilities and limitations to evaluating the validity of multiply imputed data. The goal of this research project is to develop novel methodology and guidelines for evaluating MI methods, and implement these in an interactive evaluation framework for multiple imputation. This framework will aid applied researchers in drawing valid inference from incomplete datasets.
